---
title: "Callbacks from C to R"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
helper <- if (file.exists("vignettes/articles/_helpers.R")) {
    "vignettes/articles/_helpers.R"
} else {
    "_helpers.R"
}
source(helper)
library(rdyncall)
```

Some C APIs do not just call functions immediately. They store a function
pointer and call it later. `ccallback()` creates a C function pointer that
enters R when the foreign code invokes it.

## A minimal callback

This callback has the C shape:

```c
int add(int x, int y);
```

```{r add-callback}
add <- ccallback("ii)i", function(x, y) x + y)
dyncall(add, "ii)i", 20L, 3L)
```

The callback signature follows the same call-signature format used by
`dyncall()`.

## Recursive callback call

A callback pointer is just another foreign function pointer. It can be passed
back into a call if the signature expects a pointer.

```{r factorial}
factorial_callback <- function(x, fun) {
    if (x < 2L) {
        1L
    } else {
        x * dyncall(fun, "ip)i", x - 1L, fun)
    }
}

factorial_ptr <- ccallback("ip)i", factorial_callback)
dyncall(factorial_ptr, "ip)i", 6L, factorial_ptr)
```

This example is small, but it shows the same lifetime rule as larger APIs: keep
`factorial_ptr` reachable in R for as long as C may call it.

## Sorting with C `qsort`

`qsort()` is a classic C API that accepts a comparison callback:

```c
void qsort(void *base, size_t nmemb, size_t size,
           int (*compar)(const void *, const void *));
```

The comparison callback receives pointers to two elements. The R callback reads
the `double` values with `unpack()`.

```{r qsort}
libc_names <- c("msvcrt", "c", "c.so.6")
libc <- new.env(parent = globalenv())
invisible(dynbind(libc_names, "qsort(pLLp)v;", envir = libc))

compare_double <- function(px, py) {
    x <- unpack(px, 0L, "d")
    y <- unpack(py, 0L, "d")
    if (x < y) {
        -1L
    } else if (x > y) {
        1L
    } else {
        0L
    }
}

set.seed(42)
x <- rnorm(12L)
expected <- sort(x)
compare_ptr <- ccallback("pp)i", compare_double)

libc$qsort(x, length(x), 8L, compare_ptr)
all.equal(x, expected)
```

The callback object must remain assigned to `compare_ptr` until `qsort()`
returns. For APIs that store callbacks beyond the registration call, store the
callback object somewhere with the same lifetime as the foreign registration.

## Immediate and stored callbacks

Callback lifetime depends on whether C calls the pointer during the current
function call or stores it for later.

| C API pattern | Example | R-side lifetime rule |
|:--|:--|:--|
| Immediate callback | `qsort()` | keep the callback object alive until the call returns |
| Stored callback | GUI, audio, event, or async APIs | keep the callback object alive until the foreign registration is removed |

`qsort()` is comparatively safe because it finishes before control returns to
R. Event loops are riskier because C may call the pointer long after the
registration function has returned.

## Error and lifetime rules

Callbacks cross from C into R, so mistakes are more severe than ordinary R
function errors.

- The callback signature must match the exact C callback type.
- Keep the callback object alive in R while foreign code may call it.
- Do not let C call a callback after its R state has been cleaned up.
- Avoid throwing errors through foreign event loops. Handle errors inside the
  callback and return a C-level failure value when the API supports one.
- Aggregate by-value callback arguments and returns are not currently supported.

## A safe registration pattern

For APIs that register callbacks, keep a small R object that owns both the
foreign handle and the callback pointer.

```{r registration-pattern, eval=FALSE}
make_registration <- function(handle) {
    state <- new.env(parent = emptyenv())
    state$handle <- handle
    state$callback <- ccallback("ip)v", function(code, userdata) {
        # Handle the event and return a C-compatible value.
        invisible(NULL)
    })

    foreign_register_callback(handle, state$callback)
    state
}
```

The important part is not the exact wrapper shape; it is that the callback
pointer remains reachable for at least as long as the C library can use it.
