---
title: "Creating DynPort files with porter"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
helper <- if (file.exists("vignettes/articles/_helpers.R")) {
    "vignettes/articles/_helpers.R"
} else {
    "_helpers.R"
}
source(helper)
library(rdyncall)
```

`rdyncall` loads DynPort metadata; it does not parse C headers by itself.
[porter](https://github.com/hongyuanjia/porter) is the companion tool used to
turn C headers into DCF `.dynport` files that `dynport()` can load.

Use this article when you want to create a DynPort file for a C library that you
use. SDL3 appears here only as a concrete example because rdyncall ships one
maintained SDL3 DynPort file.

## Workflow overview

The generation boundary is:

1. Install or locate the C library headers and the runtime library.
2. Write a small umbrella header that includes the public header you want to
   bind.
3. Run [porter](https://github.com/hongyuanjia/porter) with the include paths
   and parse boundary for that library.
4. Set DynPort metadata such as `Package`, `Version`, and `Library`.
5. Write a DCF `.dynport` file and review the generated diff.
6. Let rdyncall load that file with `dynport()`.

Only step 6 happens at rdyncall runtime. Header parsing and metadata generation
are development-time steps.

## Choose the inputs

Before running [porter](https://github.com/hongyuanjia/porter), decide these
inputs for the library you want to bind:

| Input | Purpose | Example for SDL3 |
|:--|:--|:--|
| Umbrella header | Keeps the parsed API surface explicit | `#include <SDL3/SDL.h>` |
| Include flags | Lets the parser find library headers | `-I/path/to/include` |
| Parse boundary | Avoids accidentally binding unrelated dependency headers | `/path/to/include/SDL3` |
| DynPort package name | Names the generated R package | `SDL3` |
| Runtime library candidates | Names that `dynfind()` can resolve on target systems | `SDL3`, `SDL3-0`, `SDL3-3` |

Small umbrella headers usually make better DynPort files than asking the parser
to follow every header installed on the system.

## Generate a DynPort file

The exact include path depends on how the target library is installed. This
script shows the pattern with SDL3 as the example:

```{r porter-script, eval=FALSE}
library(porter)

sdl3_header <- "/path/to/include/SDL3/SDL.h"
sdl3_include <- dirname(dirname(sdl3_header))

umbrella <- tempfile(fileext = ".h")
writeLines("#include <SDL3/SDL.h>", umbrella)

sdl3 <- port(
    umbrella,
    limit = dirname(sdl3_header),
    cflags = paste0("-I", sdl3_include)
)

sdl3 <- port_set(
    sdl3,
    Package = "SDL3",
    Version = "3.4.10",
    Library = c("SDL3", "SDL3-0", "SDL3-3")
)

port_write(sdl3, "inst/dynports/SDL3.dynport")
```

For another library, replace the header path, umbrella include, `Package`,
`Version`, and `Library` values with the values for that library.

Keep generated output reviewed in git. A changed header,
[porter](https://github.com/hongyuanjia/porter) release, or upstream library
version can legitimately change function, constant, enum, and aggregate
metadata.

## Inspect an existing DynPort file

The installed rdyncall package contains one bundled DynPort file for SDL3:

```{r bundled-file}
sdl3_portfile <- system.file("dynports", "SDL3.dynport",
                            package = "rdyncall", mustWork = TRUE)
basename(sdl3_portfile)
```

The file is plain text DCF metadata. The header shows the package identity and
library candidates:

```{r bundled-header}
head(readLines(sdl3_portfile), 12L)
```

The `Library` entries are candidates, not a guarantee that SDL3 is installed on
the current machine. `dynport()` resolves them when loading the generated
package.

## Validate on the rdyncall side

After generating a file, validate the rdyncall side separately from the
header-generation step:

```{r validate-shape, eval=FALSE}
portfile <- "inst/dynports/YourLibrary.dynport"
lib <- tempfile("rdyncall-dynport-lib")

generated <- dynport(portfile = portfile, package = "dyn.YourLibraryCheck",
                     lib = lib, rebuild = TRUE, quiet = TRUE)
generated
```

If the target library is installed, call a non-GUI scalar function first. For
SDL3, `SDL_GetPlatform()` is a good smoke test because it does not open a window
or start an event loop.

## What belongs where

| Concern | Home |
|:--|:--|
| Header parsing and metadata generation | [porter](https://github.com/hongyuanjia/porter) |
| Reviewed DCF file | Your package or `inst/dynports/*.dynport` |
| Loading generated packages | `dynport()` |
| Real function calls | generated package wrappers |
| GUI demos and event loops | separate demos or articles |

The rdyncall package intentionally keeps the bundled DynPort surface small. SDL3
is the maintained example. Additional libraries should be generated from their
current headers rather than copied from historical DynPort files.

## Next steps

- Use [dynbind and DynPort bindings](dynbind-dynport.html) for the rdyncall
  loading layer.
- Use [SDL3 non-GUI probing](sdl3-non-gui.html) for a CI-safe call through the
  bundled SDL3 example.
- Use [troubleshooting](troubleshooting.html) when a generated DynPort package
  cannot find its library or reports unresolved symbols.
