---
title: "dynbind and DynPort bindings"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
helper <- if (file.exists("vignettes/articles/_helpers.R")) {
    "vignettes/articles/_helpers.R"
} else {
    "_helpers.R"
}
source(helper)
library(rdyncall)
```

Direct `dyncall()` calls are useful for one-off experiments. Larger bindings
usually need names, wrappers, constants, types, and repeatable loading. That is
where `dynbind()` and `dynport()` fit.

Use this article when you already know several function signatures and want to
move beyond address-by-address calls. It shows when to keep a small binding in
an environment with `dynbind()`, and when to put binding metadata in a DynPort
file so `dynport()` can generate and load an R package.

The three layers have different jobs:

| Layer | Input | Output | Best use |
|:--|:--|:--|:--|
| `dyncall()` | a function pointer and one call signature | one foreign call | probes and one-off calls |
| `dynbind()` | library candidates and hand-written signatures | R wrappers in an environment | small, explicit bindings |
| `dynport()` | a DCF `.dynport` file | a generated R package | repeatable bindings from metadata |
| [porter](https://github.com/hongyuanjia/porter) | C headers | a `.dynport` file | generating metadata outside rdyncall |

[porter](https://github.com/hongyuanjia/porter) is not required at run time by
rdyncall. It is the tool used to generate DynPort files from headers before
`dynport()` loads those files.

## Bind a small function set

`dynbind()` takes a library name or handle and a library signature. It installs
thin R wrappers in an environment.

```{r dynbind-math}
math_names <- c("msvcrt", "m", "m.so.6")
math <- new.env(parent = globalenv())
info <- dynbind(
    math_names,
    paste(
        "sqrt(d)d",
        "cos(d)d",
        "sin(d)d",
        sep = ";"
    ),
    envir = math
)

c(
    sqrt = math$sqrt(49),
    cos = math$cos(0),
    sin = math$sin(pi / 2)
)
```

The generated wrapper still calls the resolved C function through rdyncall. The
wrapper just hides the address and signature from the user-facing call.

## Inspect unresolved symbols

The return value reports the library handle and unresolved symbols. This makes
it possible to fail early when a platform lacks a function you require.

```{r unresolved}
str(info$unresolved.symbols)
```

For exploratory bindings, unresolved optional symbols can be reported to the
user. For required bindings, fail before exposing partially working wrappers.

## Library discovery

Character library names that look like file paths are loaded directly.
Other character values are resolved with `dynfind()`.

```{r dynfind}
libc_names <- c("msvcrt", "c", "c.so.6")
libc_handle <- dynfind(libc_names)
is.nullptr(libc_handle)
```

Use a vector of candidate names for cross-platform bindings. For example,
Windows may use `msvcrt`, while Unix-like systems commonly expose C runtime
symbols through `c` or `c.so.6`.

## From bindings to generated packages

`dynport()` reads a DCF `.dynport` file and generates a real R package in an
rdyncall-managed library. The generated package namespace is populated from the
metadata in the DynPort file.

The default DynPort library is:

```{r dynport-lib}
dynport_lib()
```

Generated package names use the option
`rdyncall.dynport.package.prefix`, whose default value is `"dyn."`, plus the
`Package` field in the DynPort file.

## DynPort file shape

A minimal DynPort file records the package name, library candidates, function
bindings, and optionally constants and aggregate types.

```text
Package: SDL3
Version: 3.4.10
Library:
    SDL3
    SDL3-0
    SDL3-3
Function:
    SDL_GetPlatform()Z;
```

The package ships one current-format DynPort file:
`inst/dynports/SDL3.dynport`. It is generated from SDL3 headers with
[porter](https://github.com/hongyuanjia/porter) and kept in the package as a
realistic, non-toy binding example.

`rdyncall` intentionally does not bundle a broad catalog of old DynPort files.
The supported in-package example is SDL3. For other libraries, generate a fresh
DCF file for the headers and library version you want to bind.

For other C libraries, generate a `.dynport` file with
[porter](https://github.com/hongyuanjia/porter) and pass it to
`dynport(portfile = ...)`. A regeneration script for SDL3 follows this shape once
the SDL3 header directory is known:

```{r creating-dynports-sdl3, eval=FALSE}
library(porter)

sdl3_header <- "/path/to/include/SDL3/SDL.h"
sdl3_include <- dirname(dirname(sdl3_header))

header <- tempfile(fileext = ".h")
writeLines("#include <SDL3/SDL.h>", header)

sdl3 <- port(header,
    limit = dirname(sdl3_header),
    cflags = paste0("-I", sdl3_include)
)
sdl3 <- port_set(sdl3,
    Package = "SDL3",
    Version = "3.4.10",
    Library = c("SDL3", "SDL3-0", "SDL3-3")
)

port_write(sdl3, "inst/dynports/SDL3.dynport")
```

## When to choose each layer

- Use `dyncall()` for a single function address or a small experiment.
- Use `dynbind()` when you know a small set of functions and want ordinary R
  wrappers in one environment.
- Use `dynport()` when the binding metadata should live in a data file and load
  as a generated R package.
- Use [porter](https://github.com/hongyuanjia/porter) when the metadata should
  come from C headers rather than from hand-written signatures.

All three layers use the same underlying signatures, so start by getting the C
declarations and signatures correct.

## Next steps

- Use [signatures](signatures.html) before writing library signatures or
  DynPort `Function` entries.
- Use [Creating DynPort files with porter](creating-dynports.html) for the
  header-to-DynPort workflow, including the SDL3 example that produces the
  bundled DynPort file.
- Use [Non-GUI demos](non-gui-demos.html) for examples that run without opening
  windows.
- Use [troubleshooting](troubleshooting.html) when generated packages fail to
  find a library, resolve a symbol, or load into the current session.
