---
title: "dynbind and DynPort bindings"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
helper <- if (file.exists("vignettes/articles/_helpers.R")) {
    "vignettes/articles/_helpers.R"
} else {
    "_helpers.R"
}
source(helper)
library(rdyncall)
```

Direct `dyncall()` calls are useful for one-off experiments. Larger bindings
usually need names, wrappers, constants, types, and repeatable loading. That is
where `dynbind()` and `dynport()` fit.

## Bind a small function set

`dynbind()` takes a library name or handle and a library signature. It installs
thin R wrappers in an environment.

```{r dynbind-math}
math <- new.env(parent = globalenv())
info <- dynbind(
    article_math_names,
    paste(
        "sqrt(d)d",
        "cos(d)d",
        "sin(d)d",
        sep = ";"
    ),
    envir = math
)
article_expect_symbols(info, "math library")

c(
    sqrt = math$sqrt(49),
    cos = math$cos(0),
    sin = math$sin(pi / 2)
)
```

The generated wrapper still calls the resolved C function through rdyncall. The
wrapper just hides the address and signature from the user-facing call.

## Inspect unresolved symbols

The return value reports the library handle and unresolved symbols. This makes
it possible to fail early when a platform lacks a function you require.

```{r unresolved}
str(info$unresolved.symbols)
```

For exploratory bindings, unresolved optional symbols can be reported to the
user. For required bindings, fail before exposing partially working wrappers.

## Library discovery

Character library names that look like file paths are loaded directly.
Other character values are resolved with `dynfind()`.

```{r dynfind}
libc_handle <- dynfind(article_libc_names)
is.nullptr(libc_handle)
```

Use a vector of candidate names for cross-platform bindings. For example,
Windows may use `msvcrt`, while Unix-like systems commonly expose C runtime
symbols through `c` or `c.so.6`.

## From bindings to generated packages

`dynport()` reads a DCF `.dynport` file and generates a real R package in an
rdyncall-managed library. The generated package namespace is populated from the
metadata in the DynPort file.

The default DynPort library is:

```{r dynport-lib}
dynport_lib()
```

Generated package names use the option
`rdyncall.dynport.package.prefix`, whose default value is `"dyn."`, plus the
`Package` field in the DynPort file.

## DynPort file shape

A minimal DynPort file records the package name, library candidates, function
bindings, and optionally constants and aggregate types.

```text
Package: SDL2
Version: 2.28.0
Library:
    SDL
    SDL-2.28
    SDL-2.28.so.0
Function:
    SDL_GetPlatform()Z;
```

The source tree includes `inst/dynports/SDL2.dynport` as a DCF-style example.
The executable example below writes a minimal temporary DynPort so it can use
modern SDL2 library names commonly seen on development machines and CI images.

## A non-GUI SDL2 probe

When `RDYNCALL_ARTICLE_EXTERNAL=true` and SDL2 is available, this article builds
and loads the generated package, then calls a non-GUI platform query. The chunk
does not open a window or enter an event loop.

```{r sdl2-dynport, eval=article_eval_external(article_sdl2_names, "SDL2")}
portfile <- tempfile("SDL2Article-", fileext = ".dynport")
writeLines(c(
    "Package: SDL2Article",
    "Version: 2.0.0",
    "Library:",
    "    SDL2",
    "    SDL2-2.0",
    "    SDL2-2.0.so.0",
    "    SDL2-2",
    "    SDL2-0",
    "Function:",
    "    SDL_GetPlatform()Z;"
), portfile)

generated <- dynport(portfile = portfile, package = "dyn.SDL2Article",
                     rebuild = TRUE, quiet = TRUE)
generated

dyn.SDL2Article::SDL_GetPlatform()
```

If SDL2 is not installed, the code remains visible in the article but is skipped
unless external article execution is explicitly requested.

## When to choose each layer

- Use `dyncall()` for a single function address or a small experiment.
- Use `dynbind()` when you know a small set of functions and want ordinary R
  wrappers in one environment.
- Use `dynport()` when the binding metadata should live in a data file and load
  as a generated R package.

All three layers use the same underlying signatures, so start by getting the C
declarations and signatures correct.
