---
title: "dynbind and DynPort bindings"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
helper <- if (file.exists("vignettes/articles/_helpers.R")) {
    "vignettes/articles/_helpers.R"
} else {
    "_helpers.R"
}
source(helper)
library(rdyncall)
```

Direct `dyncall()` calls are useful for one-off experiments. Larger bindings
usually need names, wrappers, constants, types, and repeatable loading. That is
where `dynbind()` and `dynport()` fit.

## Bind a small function set

`dynbind()` takes a library name or handle and a library signature. It installs
thin R wrappers in an environment.

```{r dynbind-math}
math_names <- c("msvcrt", "m", "m.so.6")
math <- new.env(parent = globalenv())
info <- dynbind(
    math_names,
    paste(
        "sqrt(d)d",
        "cos(d)d",
        "sin(d)d",
        sep = ";"
    ),
    envir = math
)
article_expect_symbols(info, "math library")

c(
    sqrt = math$sqrt(49),
    cos = math$cos(0),
    sin = math$sin(pi / 2)
)
```

The generated wrapper still calls the resolved C function through rdyncall. The
wrapper just hides the address and signature from the user-facing call.

## Inspect unresolved symbols

The return value reports the library handle and unresolved symbols. This makes
it possible to fail early when a platform lacks a function you require.

```{r unresolved}
str(info$unresolved.symbols)
```

For exploratory bindings, unresolved optional symbols can be reported to the
user. For required bindings, fail before exposing partially working wrappers.

## Library discovery

Character library names that look like file paths are loaded directly.
Other character values are resolved with `dynfind()`.

```{r dynfind}
libc_names <- c("msvcrt", "c", "c.so.6")
libc_handle <- dynfind(libc_names)
is.nullptr(libc_handle)
```

Use a vector of candidate names for cross-platform bindings. For example,
Windows may use `msvcrt`, while Unix-like systems commonly expose C runtime
symbols through `c` or `c.so.6`.

## From bindings to generated packages

`dynport()` reads a DCF `.dynport` file and generates a real R package in an
rdyncall-managed library. The generated package namespace is populated from the
metadata in the DynPort file.

The default DynPort library is:

```{r dynport-lib}
dynport_lib()
```

Generated package names use the option
`rdyncall.dynport.package.prefix`, whose default value is `"dyn."`, plus the
`Package` field in the DynPort file.

## DynPort file shape

A minimal DynPort file records the package name, library candidates, function
bindings, and optionally constants and aggregate types.

```text
Package: SDL3
Version: 3.4.10
Library:
    SDL3
    SDL3-0
    SDL3-3
Function:
    SDL_GetPlatform()Z;
```

The package ships one current-format DynPort file:
`inst/dynports/SDL3.dynport`. It is generated from SDL3 headers with
[porter](https://github.com/hongyuanjia/porter) and kept in the package as a
realistic, non-toy binding example.

For other C libraries, generate a `.dynport` file with porter and pass it to
`dynport(portfile = ...)`. A regeneration script for SDL3 follows this shape
once the SDL3 header directory is known:

```{r porter-sdl3, eval=FALSE}
library(porter)

sdl3_header <- "/path/to/include/SDL3/SDL.h"
sdl3_include <- dirname(dirname(sdl3_header))

header <- tempfile(fileext = ".h")
writeLines("#include <SDL3/SDL.h>", header)

sdl3 <- port(header,
    limit = dirname(sdl3_header),
    cflags = paste0("-I", sdl3_include)
)
sdl3 <- port_set(sdl3,
    Package = "SDL3",
    Version = "3.4.10",
    Library = c("SDL3", "SDL3-0", "SDL3-3")
)

port_write(sdl3, "inst/dynports/SDL3.dynport")
```

## A non-GUI SDL3 probe

When `RDYNCALL_ARTICLE_EXTERNAL=true` and SDL3 is available, this article builds
and loads a generated package from the bundled DynPort, then calls a non-GUI
platform query. The chunk does not open a window or enter an event loop.

```{r sdl3-names}
sdl3_names <- c("SDL3", "SDL3-0", "SDL3-3")
```

During normal local rendering, `RDYNCALL_ARTICLE_EXTERNAL` can stay unset or
`false`; the SDL3 code is shown but skipped. The pkgdown workflow sets it to
`true`, installs SDL3 system libraries, and therefore executes this non-GUI
probe in CI.

```{r external-control, eval=FALSE}
Sys.setenv(RDYNCALL_ARTICLE_EXTERNAL = "true")
```

```{r sdl3-dynport, eval=article_eval_external(sdl3_names, "SDL3")}
portfile <- system.file("dynports", "SDL3.dynport",
                        package = "rdyncall", mustWork = TRUE)
lib <- tempfile("rdyncall-dynport-lib")

generated <- dynport(portfile = portfile, package = "dyn.SDL3Article",
                     lib = lib, rebuild = TRUE, quiet = TRUE)
generated

getExportedValue(generated, "SDL_GetPlatform")()
```

If SDL3 is not installed, the code remains visible in the article but is skipped
unless external article execution is explicitly requested.

## When to choose each layer

- Use `dyncall()` for a single function address or a small experiment.
- Use `dynbind()` when you know a small set of functions and want ordinary R
  wrappers in one environment.
- Use `dynport()` when the binding metadata should live in a data file and load
  as a generated R package.

All three layers use the same underlying signatures, so start by getting the C
declarations and signatures correct.
