---
title: "Getting started with rdyncall"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
helper <- if (file.exists("vignettes/articles/_helpers.R")) {
    "vignettes/articles/_helpers.R"
} else {
    "_helpers.R"
}
source(helper)
library(rdyncall)
```

`rdyncall` is a low-level Foreign Function Interface (FFI) for R. It lets you
load shared C libraries, resolve symbols, and call C functions directly from R
when you know the C declaration you want to bind.

It is useful for exploration, prototypes, dynamic bindings, and small direct
interfaces. For stable high-level R packages, a compiled `.Call` wrapper may
still be the better long-term interface.

## A first direct call

The basic workflow has three steps:

1. Find or load a shared library.
2. Resolve a function symbol.
3. Call the address with a signature that matches the C function type.

The C math function is declared as:

```c
double sqrt(double x);
```

The corresponding rdyncall call signature is `"d)d"`: one `double` argument,
then `)`, then a `double` return value.

```{r sqrt}
mathlib <- dynfind(article_math_names)
sqrt_addr <- dynsym(mathlib, "sqrt")
dyncall(sqrt_addr, "d)d", 144)
```

If the signature does not match the real C function type, the process can
crash. Keep the C declaration beside the R binding while you develop.

## Wrap the same idea

`dynbind()` creates thin R wrappers for one or more functions in a library.
The wrapper still uses the same signature internally, but the call site becomes
ordinary R code.

```{r dynbind-sqrt}
math <- new.env(parent = globalenv())
info <- dynbind(article_math_names, "sqrt(d)d;", envir = math)
article_expect_symbols(info, "math library")

math$sqrt(625)
```

## Call an R function through a C callback pointer

`ccallback()` turns an R function into a C function pointer. The signature
describes the callback function type.

```{r callback-add}
add <- ccallback("ii)i", function(x, y) x + y)
dyncall(add, "ii)i", 20L, 3L)
```

Keep an R reference to callback objects for as long as foreign code may call
them. If a callback is garbage-collected while C still holds its pointer, the
next C call can crash R.

## Describe and use a C struct

C aggregate types are registered with compact structure signatures. This C type:

```c
struct Rect {
  short x;
  short y;
  unsigned short w;
  unsigned short h;
};
```

can be registered and used from R:

```{r rect}
cstruct("ArticleRect{ssSS}x y w h;")

rect <- cdata(ArticleRect)
rect$x <- 40L
rect$y <- 60L
rect$w <- 10L
rect$h <- 15L

rect$w * rect$h
```

`cdata()` returns a raw-backed object with struct metadata. Field access reads
and writes the raw bytes according to the registered layout.

## What to learn next

- Use [signatures](signatures.html) to translate C declarations into rdyncall
  signatures.
- Use [structs, unions, and memory](structs-unions-memory.html) for aggregate
  layouts and raw memory access.
- Use [callbacks](callbacks.html) when a C API stores and later calls an R
  function pointer.
- Use [dynbind and dynport](dynbind-dynport.html) for larger bindings.
