---
title: "GUI demos"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rdyncall)
```

`rdyncall` can call real C GUI and graphics libraries directly from R, including
interactive SDL3 windows, audio/visual examples, OpenGL rendering, and raylib
drawing loops.

## SDL3 Snake

The SDL3 Snake demo opens a window, polls keyboard input, updates game state in
R, and renders each frame through SDL3 calls made with `dynbind()`.

<video controls muted loop playsinline width="100%" poster="https://github.com/hongyuanjia/rdyncall/releases/download/docs-media/rdyncall-sdl3-snake-poster.png">
  <source src="https://github.com/hongyuanjia/rdyncall/releases/download/docs-media/rdyncall-sdl3-snake.mp4" type="video/mp4">
</video>

```{r sdl-snake, eval=FALSE}
demo("SDL", package = "rdyncall", ask = FALSE)
```

The demo uses `cstruct()` for `SDL_FRect`, `dynbind()` for the SDL3 renderer and
event functions, a raw buffer for `SDL_PollEvent()`, and keyboard state returned
from `SDL_GetKeyboardState()`.

## SDL3 audio

The SDL3 audio demo combines a small visual window with planar audio buffers.
It prepares low-level audio structs and feeds sample data into an SDL3 audio
stream from R.

<!-- Replace with a screenshot or video after uploading media:
<video controls muted loop playsinline width="100%" poster="https://github.com/hongyuanjia/rdyncall/releases/download/docs-media/rdyncall-sdl3-audio-poster.png">
  <source src="https://github.com/hongyuanjia/rdyncall/releases/download/docs-media/rdyncall-sdl3-audio.mp4" type="video/mp4">
</video>
-->

```{r sdl-audio, eval=FALSE}
demo("SDL_audio", package = "rdyncall", ask = FALSE)
```

This example shows how rdyncall can pass aggregate pointers, raw sample buffers,
and runtime-managed native resources through a C API that owns audio playback.

## SDL3/OpenGL raster

The SDL3/OpenGL raster demo opens an SDL3 window with an OpenGL context and
renders a Mandelbrot raster. It is a compact example of using rdyncall for a
graphics pipeline rather than only individual scalar function calls.

<!-- Replace with a screenshot or video after uploading media:
<video controls muted loop playsinline width="100%" poster="https://github.com/hongyuanjia/rdyncall/releases/download/docs-media/rdyncall-sdl3-raster-poster.png">
  <source src="https://github.com/hongyuanjia/rdyncall/releases/download/docs-media/rdyncall-sdl3-raster.mp4" type="video/mp4">
</video>
-->

```{r sdl-raster, eval=FALSE}
demo("SDL_raster", package = "rdyncall", ask = FALSE)
```

The raster demo exercises SDL3 window creation, OpenGL context setup, native
buffer uploads, and a timed render loop controlled from R.

## raylib recursive tree

The raylib recursive tree demo draws animated branch geometry with raylib. The R
side generates branch endpoints and passes raylib aggregate values such as
`Vector2` and `Color` by value.

<!-- Replace with a screenshot or video after uploading media:
<video controls muted loop playsinline width="100%" poster="https://github.com/hongyuanjia/rdyncall/releases/download/docs-media/rdyncall-raylib-tree-poster.png">
  <source src="https://github.com/hongyuanjia/rdyncall/releases/download/docs-media/rdyncall-raylib-tree.mp4" type="video/mp4">
</video>
-->

```{r raylib-tree, eval=FALSE}
demo("raylib", package = "rdyncall", ask = FALSE)
```

This demo highlights aggregate-by-value calls, generated geometry in R, and a
foreign drawing loop managed through raylib.

## What these demos exercise

| Demo | Native library | rdyncall features |
|:--|:--|:--|
| SDL3 Snake | SDL3 renderer and events | `dynbind()`, `cstruct()`, raw event buffer, keyboard state, event loop |
| SDL3 audio | SDL3 audio and renderer | aggregate pointers, raw audio buffers, native resource cleanup |
| SDL3/OpenGL raster | SDL3 and OpenGL | shared-library discovery, OpenGL context setup, raster buffer upload, timed rendering |
| raylib recursive tree | raylib | aggregate by value, generated R geometry, drawing loop, downloaded native library |

Together, the demos show rdyncall operating at the boundary where R code owns the
application logic while C libraries own windows, renderers, audio streams, and
graphics contexts.

## Next steps

- Use [SDL3 non-GUI probing](sdl3-non-gui.html) for a CI-friendly SDL3 binding
  check.
- Use [dynbind and DynPort bindings](dynbind-dynport.html) to understand the
  wrapper layer used by the demos.
- Use [structs, unions, and memory](structs-unions-memory.html) for aggregate
  values such as `SDL_FRect`, `Color`, and `Vector2`.
- Use [FFI safety boundaries](ffi-safety.html) before adapting these patterns to
  APIs that store pointers, register callbacks, or own native resources.
