---
title: "Non-GUI demos"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(rdyncall)
```

`rdyncall` can call real C libraries without opening windows: parse XML, sort R
vectors with C algorithms, solve optimization problems, and move bytes through
low-level I/O APIs.

## XML parsing with libxml2

The libxml2 demo binds the streaming reader API directly with `dynbind()`. R
owns the XML text and result assembly; libxml2 owns the parser state.

```{r libxml2-demo, eval=FALSE}
demo("libxml2", package = "rdyncall", ask = FALSE)
```

The direct binding uses `xmlReaderForMemory()` to create a reader,
`xmlTextReaderRead()` to iterate nodes, and `xmlTextReaderConstValue()` to read
text values. The reader pointer is released with `xmlFreeTextReader()`.

For comparison, the same task with the high-level `xml2` package looks like:

```{r xml2-comparison, eval=FALSE}
xml_text <- "<root><message>rdyncall libxml2 demo</message></root>"
doc <- xml2::read_xml(xml_text)
xml2::xml_text(xml2::xml_find_first(doc, ".//message"))
```

The `xml2` version is the right interface for everyday XML work. The rdyncall
version shows what is happening one layer lower: shared-library discovery,
symbol binding, native pointer lifetime, and direct calls into libxml2.

## Sorting through C qsort

The `qsort` demo passes an R numeric vector to C and gives C a comparator
implemented as an R function through `ccallback()`.

```{r qsort-demo, eval=FALSE}
demo("qsort", package = "rdyncall", ask = FALSE)
```

The result matches `sort(x)`, but the sorting loop is the C standard library
calling back into R for each comparison.

## Solving a linear program with GLPK

The GLPK demo creates a native optimization problem, loads a sparse constraint
matrix, runs the simplex solver, and reads the objective value and primal
solution back into R.

```{r glpk-demo, eval=FALSE}
demo("glpk", package = "rdyncall", ask = FALSE)
```

The important FFI pattern is ownership: `glp_create_prob()` returns a problem
pointer, R configures that object through GLPK calls, and `glp_delete_prob()`
releases the native object when the demo exits.

## Low-level stdio with raw vectors

The stdio demo shows how a C API can read and write bytes while R manages the
data as raw vectors.

```{r stdio-demo, eval=FALSE}
demo("stdio", package = "rdyncall", ask = FALSE)
```

This style is useful for understanding APIs built around `FILE*`, pointer
arguments, byte counts, and explicit cleanup.

## More small demos

| Demo | What it shows |
|:--|:--|
| `sqrt` | Minimal dynamic library lookup and scalar call |
| `callbacks` | Creating a C-callable function pointer from an R function |
| `factorial` | Recursive callback-driven control flow |
| `R_ShowMessage` | Calling an R C API function from R through rdyncall |

These smaller demos are useful when learning one FFI concept at a time before
moving to larger library bindings.

## Next steps

- Use [signatures](signatures.html) to translate C prototypes into rdyncall
  call signatures.
- Use [callbacks](callbacks.html) before passing R functions to C APIs.
- Use [structs, unions, and memory](structs-unions-memory.html) for pointer,
  raw-vector, and aggregate data patterns.
- Use [FFI safety boundaries](ffi-safety.html) before adapting these examples to
  ownership-sensitive libraries.
