---
title: "Porter and SDL3 DynPort"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
helper <- if (file.exists("vignettes/articles/_helpers.R")) {
    "vignettes/articles/_helpers.R"
} else {
    "_helpers.R"
}
source(helper)
library(rdyncall)
```

`rdyncall` loads DynPort metadata; it does not parse C headers by itself.
[porter](https://github.com/hongyuanjia/porter) is the companion tool used to
turn C headers into DCF `.dynport` files. The bundled SDL3 DynPort is the
package's current real-world example of that workflow.

Use this article when you want to regenerate `inst/dynports/SDL3.dynport` from
SDL3 headers, or when you want to follow the same pattern for another C
library.

## Workflow overview

The generation boundary is:

1. Install or locate the C library headers.
2. Run [porter](https://github.com/hongyuanjia/porter) against a small umbrella
   header.
3. Set DynPort metadata such as `Package`, `Version`, and `Library`.
4. Write a DCF `.dynport` file.
5. Let rdyncall load that file with `dynport()`.

Only step 5 happens at rdyncall runtime.

## The bundled SDL3 DynPort

The installed package contains one bundled DynPort file for SDL3:

```{r bundled-file}
sdl3_portfile <- system.file("dynports", "SDL3.dynport",
                            package = "rdyncall", mustWork = TRUE)
basename(sdl3_portfile)
```

The file is plain text DCF metadata. The header shows the package identity and
library candidates:

```{r bundled-header}
head(readLines(sdl3_portfile), 12L)
```

The `Library` entries are candidates, not a guarantee that SDL3 is installed on
the current machine. `dynport()` resolves them when loading the generated
package.

## Regenerate from SDL3 headers

The exact include path depends on how SDL3 is installed. The shape of the
generation script is:

```{r porter-script, eval=FALSE}
library(porter)

sdl3_header <- "/path/to/include/SDL3/SDL.h"
sdl3_include <- dirname(dirname(sdl3_header))

umbrella <- tempfile(fileext = ".h")
writeLines("#include <SDL3/SDL.h>", umbrella)

sdl3 <- port(
    umbrella,
    limit = dirname(sdl3_header),
    cflags = paste0("-I", sdl3_include)
)

sdl3 <- port_set(
    sdl3,
    Package = "SDL3",
    Version = "3.4.10",
    Library = c("SDL3", "SDL3-0", "SDL3-3")
)

port_write(sdl3, "inst/dynports/SDL3.dynport")
```

Keep generated output reviewed in git. A changed header,
[porter](https://github.com/hongyuanjia/porter) release, or SDL3 version can
legitimately change function, constant, enum, and aggregate metadata.

## Validate on the rdyncall side

After regenerating the file, validate the rdyncall side separately from the
header-generation step:

```{r validate-shape, eval=FALSE}
portfile <- "inst/dynports/SDL3.dynport"
lib <- tempfile("rdyncall-dynport-lib")

generated <- dynport(portfile = portfile, package = "dyn.SDL3Check",
                     lib = lib, rebuild = TRUE, quiet = TRUE)
generated
```

If SDL3 is installed, call a non-GUI function such as `SDL_GetPlatform()` to
prove the generated package can resolve and call at least one real symbol.

## What belongs where

| Concern | Home |
|:--|:--|
| Header parsing and metadata generation | [porter](https://github.com/hongyuanjia/porter) |
| Bundled DCF file | `inst/dynports/SDL3.dynport` |
| Loading generated packages | `dynport()` |
| Real function calls | generated package wrappers |
| GUI demos and event loops | separate demos or articles |

The rdyncall package intentionally keeps the bundled DynPort surface small. SDL3
is the maintained example. Additional libraries should be generated from their
current headers rather than copied from historical DynPort files.

## Next steps

- Use [dynbind and DynPort bindings](dynbind-dynport.html) for the rdyncall
  loading layer.
- Use [SDL3 non-GUI probing](sdl3-non-gui.html) for a CI-safe call through the
  generated SDL3 package.
- Use [troubleshooting](troubleshooting.html) when a generated DynPort package
  cannot find SDL3 or reports unresolved symbols.
