---
title: "SDL3 non-GUI probing"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
helper <- if (file.exists("vignettes/articles/_helpers.R")) {
    "vignettes/articles/_helpers.R"
} else {
    "_helpers.R"
}
source(helper)
library(rdyncall)
```

SDL3 is a practical external-library target because it is cross-platform and
has both simple query functions and more complex GUI/event APIs. This article
stays on the safe side: it loads the bundled SDL3 DynPort and calls only
non-GUI query functions.

Use this article when you want to validate the generated SDL3 package in
pkgdown or CI without opening a window, initializing video, or entering an event
loop.

## External execution switch

Local article builds skip SDL3 calls by default. To force the probe, set:

```{r external-control, eval=FALSE}
Sys.setenv(RDYNCALL_ARTICLE_EXTERNAL = "true")
```

The pkgdown workflow sets this variable and installs SDL3, so the external
chunks run in CI. If the variable is true but SDL3 cannot be found, the helper
fails early instead of silently skipping the probe.

```{r sdl3-names}
sdl3_names <- c("SDL3", "SDL3-0", "SDL3-3")
article_library_available(sdl3_names)
```

## Generate and load the package

`dynport()` reads the bundled DCF file, generates a small R package in a
temporary library, and loads that package into the current session.

```{r load-sdl3, eval=article_eval_external(sdl3_names, "SDL3")}
portfile <- system.file("dynports", "SDL3.dynport",
                        package = "rdyncall", mustWork = TRUE)
lib <- tempfile("rdyncall-sdl3-lib")

generated <- dynport(portfile = portfile, package = "dyn.SDL3Probe",
                     lib = lib, rebuild = TRUE, quiet = TRUE)
generated
```

The generated package name is explicit so repeated article builds do not depend
on a previously generated default package.

## Call non-GUI SDL3 functions

The following calls query process and platform information. They do not open
windows or run an event loop.

```{r sdl3-platform, eval=article_eval_external(sdl3_names, "SDL3")}
sdl3_platform <- getExportedValue(generated, "SDL_GetPlatform")()
sdl3_revision <- getExportedValue(generated, "SDL_GetRevision")()
sdl3_version <- getExportedValue(generated, "SDL_GetVersion")()
sdl3_cpus <- getExportedValue(generated, "SDL_GetNumLogicalCPUCores")()

list(
    platform = sdl3_platform,
    revision = sdl3_revision,
    version = sdl3_version,
    logical_cpus = sdl3_cpus
)
```

This is a useful smoke test for the full stack: DynPort metadata, generated R
package, library discovery, symbol resolution, signatures, and one real foreign
call.

## Why not open a window here?

Window creation and event loops are valuable demos, but they add platform,
display-server, and lifecycle concerns that do not belong in a first generated
binding smoke test. Keep those examples separate so CI can validate the binding
without GUI state.

## Next steps

- Use [Porter and SDL3 DynPort](porter-sdl3.html) to regenerate the bundled
  metadata from SDL3 headers.
- Use [dynbind and DynPort bindings](dynbind-dynport.html) to understand how
  generated packages relate to hand-written wrappers.
- Use [callbacks](callbacks.html) before binding SDL3 APIs that store function
  pointers.
- Use [FFI safety](ffi-safety.html) before moving from query functions to GUI,
  audio, or event-loop APIs.
