---
title: "Structs, unions, and memory"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
helper <- if (file.exists("vignettes/articles/_helpers.R")) {
    "vignettes/articles/_helpers.R"
} else {
    "_helpers.R"
}
source(helper)
library(rdyncall)
```

Many C APIs exchange data through structs, unions, arrays, and pointers.
`rdyncall` models those layouts at run time so R code can read and write the
same memory shape that C expects.

Use this article when a C function takes a struct, fills an output buffer,
returns a pointer to memory, or expects fields with C-specific alignment,
packing, arrays, bitfields, or unions. The goal is to choose the highest-level
rdyncall memory interface that still matches the C API exactly.

## Raw memory with `pack()` and `unpack()`

The lowest-level tools are `pack()` and `unpack()`. They write and read C values
inside raw vectors or memory referenced by external pointers.

```{r pack-unpack}
buf <- raw(8)
pack(buf, 0, "f", 1.5)
pack(buf, 4, "f", 2.25)

c(
    first = unpack(buf, 0, "f"),
    second = unpack(buf, 4, "f")
)
```

Offsets are byte offsets starting at 0.

## Choosing a memory interface

Use the highest-level interface that still matches the C API you are binding.

| Need | Use |
|:--|:--|
| Read or write a named aggregate field | `$` and `$<-` on a `cdata()` object |
| Fill an output buffer or scalar field by byte offset | `pack()` |
| Read a value from a pointer or raw buffer by byte offset | `unpack()` |
| Treat existing raw memory as a registered aggregate | `as.ctype()` |
| Pass an aggregate by value | `cdata()` with a registered `cstruct()` or `cunion()` |

## Register a struct

This C type:

```c
struct Rect {
  short x;
  short y;
  unsigned short w;
  unsigned short h;
};
```

is represented by one structure signature:

```{r rect}
cstruct("DocRect{ssSS}x y w h;")
DocRect

rect <- cdata(DocRect)
rect$x <- 10L
rect$y <- 20L
rect$w <- 200L
rect$h <- 100L

rect
```

The object returned by `cdata()` is a raw vector with class metadata. Field
access is translated into byte-level reads and writes.

## Fixed-size array fields

Array lengths are written after the field type.

```{r arrays}
cstruct("DocColor{C[4]}rgba;")

color <- cdata(DocColor)
color$rgba <- c(255L, 128L, 0L, 255L)
color$rgba
```

The whole array field is assigned and read as an R vector.

## Nested aggregates

Struct fields can contain other registered aggregate types, including fixed-size
arrays of nested structs.

```{r nested}
cstruct("DocVec2{ff}x y;")
cstruct("DocSegment{<DocVec2>[2]}points;")

a <- cdata(DocVec2)
a$x <- 1.25
a$y <- 2.5

b <- cdata(DocVec2)
b$x <- 3.5
b$y <- 4.75

segment <- cdata(DocSegment)
segment$points <- list(a, b)
segment$points
```

Nested structs are still raw-backed values. This matters when they are passed by
value to C functions on supported platforms.

## Bitfields

Bitfields keep their ordinary integer storage type in the type list and put the
bit width in the field-name list.

```{r bitfields}
cstruct("DocBits{IIII}enabled:1 mode:3 :4 code:8;")

bits <- cdata(DocBits)
bits$enabled <- 1L
bits$mode <- 5L
bits$code <- 171L

c(
    enabled = bits$enabled,
    mode = bits$mode,
    code = bits$code
)
```

Unnamed bitfields such as `:4` reserve padding bits. A zero-width unnamed
bitfield, written `:0`, aligns the next bitfield to a new storage unit.

## Packed and aligned layouts

Layout directives appear after the field list.

```{r packed}
cstruct("DocPacked{Cd}tag value @packed;")
cstruct("DocPack4{Cd}tag value @pack(4);")
cstruct("DocAligned{C}tag @align(8);")

c(
    packed_size = DocPacked$size,
    packed_align = DocPacked$align,
    pack4_align = DocPack4$align,
    aligned_align = DocAligned$align
)
```

`@packed` is equivalent to `@pack(1)`. `@pack(n)` caps member alignment at `n`,
and `@align(n)` raises the final aggregate alignment to at least `n`.

## Unions

Unions use `|` instead of `{` after the type name.

```{r union}
cunion("DocValue|iC[4]}i bytes;")

value <- cdata(DocValue)
value$i <- 16909060L
value$bytes
```

All union fields share the same storage. Writing one field changes what another
field reads from the same bytes.

## Safety notes

- Register the aggregate layout before using it in `dyncall()`.
- Keep the C definition beside the R signature while developing.
- Be careful with platform-dependent layout, especially packing and bitfields.
- Treat external pointers as borrowed memory unless the C API explicitly says
  that R owns or must free the pointer.

## Next steps

- Use [signatures](signatures.html) to connect these memory layouts to function
  call signatures.
- Use [callbacks](callbacks.html) when a struct or pointer is passed into an R
  callback from C.
- Use [FFI safety](ffi-safety.html) before passing ownership-sensitive pointers
  or memory allocated by a foreign library.
- Use [troubleshooting](troubleshooting.html) when field values look shifted,
  truncated, or platform-dependent.
