---
title: "Troubleshooting rdyncall bindings"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
helper <- if (file.exists("vignettes/articles/_helpers.R")) {
    "vignettes/articles/_helpers.R"
} else {
    "_helpers.R"
}
source(helper)
library(rdyncall)
```

Foreign-function bugs usually fail at one of five boundaries: library
discovery, symbol lookup, signature translation, memory layout, or callback
lifetime. Work from the outside in and prove each boundary before adding the
next one.

## Library discovery

Start by proving that rdyncall can find and load the shared library. Use several
candidate names for cross-platform code.

```{r library-discovery}
math_names <- c("msvcrt", "m", "m.so.6")
mathlib <- dynfind(math_names)
is.nullptr(mathlib)
```

When this returns `TRUE`, inspect the exact candidates rdyncall tried:

```{r dynfind-explain}
dynfind_explain(math_names)
```

If no candidate loads, try a full path with `dynload()`, check whether the
library is installed for the current architecture, and remember that names
differ by platform and package manager.

## Windows DLL discovery

Windows failures are often caused by a DLL existing on disk while one of its
transitive DLL dependencies is missing from the loader search path. A direct
`dynload("C:/path/to/foo.dll")` can still fail when `foo.dll` depends on another
DLL that Windows cannot find.

Use `dynfind_explain()` first. It reports package-manager candidates from the R
runtime, Scoop, MSYS2, vcpkg, and conda, whether the file exists, and whether it
actually loaded.

```{r windows-dll-discovery, eval=.Platform$OS.type == "windows"}
dynfind_explain("SDL3")
```

The common fixes are:

| Installation source | Directory rdyncall checks |
|:--|:--|
| Scoop | `SCOOP/apps/<name>/current/bin` and related app directories |
| MSYS2 | `MINGW_PREFIX/bin`, `MSYSTEM_PREFIX/bin`, and common `C:/msys64` prefixes |
| vcpkg | `VCPKG_ROOT/installed/<triplet>/bin` |
| conda | `CONDA_PREFIX/Library/bin` and `CONDA_PREFIX/bin` |

For transitive DLL failures, put the dependency DLLs beside the primary DLL or
add their directory to `PATH` before starting R. After changing `PATH`, restart
R so the process has the updated loader environment.

## Symbol lookup

Once the library is loaded, resolve a known function and check the pointer.

```{r symbol-lookup}
sqrt_addr <- dynsym(mathlib, "sqrt")
is.nullptr(sqrt_addr)
```

If a library loads but a symbol is missing, verify the exact exported symbol
name with platform tools such as `nm`, `otool`, `dumpbin`, or `objdump`. C++
APIs may export mangled names unless the header uses `extern "C"`.

## Signature symptoms

Wrong signatures are the most dangerous class of error. They can return
nonsense, corrupt memory, or crash the R session.

| Symptom | First place to check |
|:--|:--|
| Correct function, wrong numeric value | scalar type width or signedness |
| Crash on return | return type or calling convention |
| Crash after several arguments | missing argument, wrong pointer type, or variadic promotion |
| String is truncated or invalid | `Z` used for data that is not a nul-terminated string |
| Vector changes unexpectedly | a pointer argument lets C mutate R memory |

Keep the C prototype beside the R signature and test with the smallest input
that exercises the binding.

## Pointer and memory issues

Treat pointers as borrowed memory unless the C API explicitly says otherwise.
Before reading or writing memory by offset, inspect the aggregate layout or
write a small raw-buffer test.

```{r memory-check}
buf <- raw(8)
pack(buf, 0, "d", 12.5)
unpack(buf, 0, "d")
```

For structs and unions, inspect size, alignment, and field offsets:

```{r layout-check}
cstruct("TroubleRect{ssSS}x y w h;")

c(
    size = TroubleRect$size,
    align = TroubleRect$align
)

TroubleRect$fields[, c("name", "type", "offset")]
```

If a field value appears in the wrong place, compare these offsets with the C
compiler's layout and check packing, alignment, bitfields, and platform-specific
type sizes.

## Callback failures

Callback problems are usually lifetime or error-boundary problems.

| Symptom | Likely cause |
|:--|:--|
| Callback works once and then crashes later | the R callback object was garbage-collected |
| Callback is called after cleanup | C still holds a pointer after R dropped state |
| Foreign event loop becomes unstable | an R error crossed the callback boundary |
| Callback receives strange values | callback signature does not match the C callback type |

Keep the `ccallback()` object reachable for as long as C may call it. For
stored callbacks, pair the registration with the C API's unregister function
and clear the R reference only after the foreign registration is gone.

## DynPort and generated packages

When `dynport()` builds a generated package but functions are missing, inspect
the returned report and the generated package name.

```{r dynport-lib}
dynport_lib()
```

Common causes are a stale `.dynport` file, a library candidate that does not
exist on the current platform, or a function listed in the metadata that is not
exported by the installed library version. Regenerate the DynPort file with
[porter](https://github.com/hongyuanjia/porter) when the headers or library
version change.

## A debugging order

1. Load the library with `dynfind()` or `dynload()`.
2. Inspect failed library loads with `dynfind_explain()`.
3. Resolve one required symbol with `dynsym()`.
4. Call the smallest scalar function first.
5. Add pointer or aggregate arguments only after the scalar call works.
6. Add callbacks only after their standalone signature has been tested.
7. Move from `dyncall()` to `dynbind()` or `dynport()` after the signature is
   proven.

## Next steps

- Use [signatures](signatures.html) to translate the C prototype you are
  debugging.
- Use [structs, unions, and memory](structs-unions-memory.html) to inspect
  layout-sensitive data.
- Use [callbacks](callbacks.html) for lifetime and error-boundary patterns.
